---
title: "BART and Overlap"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(ggpubr)
library(tidyverse)
library(shiny)
library(bartCause)
library(ggforce)
library(gridExtra)
library(cowplot)
library(ggthemes) 
library(plotly)
```

```{r, echo=FALSE}
plot_scatterplot <- function(df){
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  linetypes <- c('Mean Difference' = 'solid', 'True SATE' = 'dashed')
  ggplot(df) + 
    geom_point( aes(x = X, y = Y, color = Z ))+ 
    xlab("X") + ylab("Observed Outcome") + 
    labs(title = "Distribution of Observed Outcomes") + 
    geom_hline(aes(yintercept = mean(Y[Z == 1]), color = '1', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y[Z == 0]), color = '0', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y1), color = '1', linetype = 'True SATE')) +
    geom_hline(aes(yintercept = mean(Y0), color = '0', linetype = 'True SATE')) +
    scale_color_manual("Treated?", values = colors) + 
    scale_linetype_manual("Linetype", values = linetypes) +
    theme_minimal(base_line_size = 0.4) + 
    theme(plot.title = element_text(size = 8), 
          legend.position="bottom",
          text=element_text(family="serif", size=8)) + 
    guides(color = guide_legend(order = 1), fill = guide_legend(order = 2))
    
}

plot_marginal_distribution <- function(df) {
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  ggplot(df, aes(x = X, fill = Z)) + 
    theme_minimal(base_line_size = 0.4) +
    geom_histogram(bins = 60, alpha = 0.4, position = "identity") +
    geom_vline(data = df[which(df$Z == 0),], aes(xintercept= mean(X), color="0"), size=1) + 
    geom_vline(data = df[which(df$Z == 1),], aes(xintercept= mean(X), color="1"), size=1) +
    scale_color_manual(name = "Treated?", values = colors) + 
    scale_fill_manual(name = "Treated?", values = colors) +
    labs(title = "Distribution of Covariate X", y = "Frequency") +
    theme(legend.position="bottom",
          plot.title = element_text(size = 8), 
          text=element_text(family="serif", size=8)) 
    
}

plot_overlap <- function(df, nooverlap = FALSE){
  p <- ggplot(data = df) + geom_histogram(bins = 60, aes(x = X, color = Z, fill = Z), alpha = 0.5)
  get_hist <- function(p) {
    d <- ggplot_build(p)$data[[1]]
    data.frame(x = d$x, xmin = d$xmin, xmax = d$xmax, y = d$y, mids = (d$xmin + d$xmax)/2)
  }
  hist = get_hist(p)
  df$Z <- as.factor(df$Z)
  if(nooverlap == TRUE){
    ggplot() + 
      geom_histogram(data = df %>% filter(Z == 1),   
                     aes( x = X, y = ..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) +
      geom_histogram(data = df %>% filter(Z == 0), 
                     aes(x = X, y = -..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) + 
      labs(title = "Distribution of Covariate X", y = "Frequency") + 
      scale_fill_manual("Treated?", values=c("blue","red")) + theme_minimal(base_line_size = 0.4) +
      theme(legend.position="bottom", plot.title = element_text(size=8),
            text=element_text(family="serif", size=8)) 
  }else{
    ggplot() + 
      geom_histogram(data = df %>% filter(Z == 1), 
                     aes( x = X, y = ..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) +
      geom_histogram(data = df %>% filter(Z == 0), 
                     aes(x = X, y = -..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) + 
      labs(title = "Distribution of Covariate X", y = "Frequency") + 
      scale_fill_manual("Treated?", values=c("blue","red")) + theme_minimal(base_line_size = 0.4) +
      theme(legend.position="bottom", plot.title = element_text(size=8),
          text=element_text(family="serif", size=8)) 
  }
}

dgp_radomized_balanced <- function (n, beta_1, seed) {
  set.seed(seed) 
  pre_quiz <- rnorm(n = n, mean = 65, sd = 3)
  beta_0 <- 10
  tau <- 5
  Z <- sample(x = c(0,1), size = n, replace = TRUE)
  Y_0 <- beta_0 + beta_1 * pre_quiz + 0 + rnorm(n) 
  Y_1 <- beta_0 + beta_1 * pre_quiz + tau + rnorm(n) 
  Y <- ifelse(Z == 1, Y_1, Y_0)
  output <- data.frame(Z, pre_quiz, Y_0, Y_1, Y)
  colnames(output) <- c( "Z", "X", "Y0", "Y1", "Y")
  return (output) 
}
```

```{r, echo=FALSE}
animation <- function(df_balanced, xaxis, yaxis){
  p <- ggplot(data = df_balanced) + geom_histogram(bins = 60, aes(x = X, color = Z, fill = Z), alpha = 0.5)
  get_hist <- function(p) {
    d <- ggplot_build(p)$data[[1]]
    data.frame(x = d$x, xmin = d$xmin, xmax = d$xmax, y = d$y, mids = (d$xmin + d$xmax)/2)
  }
  hist = get_hist(p)
  hist_group0 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  hist_group1 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  
  idx_group0 <- which(df_balanced$Z == 0)
  idx_group1 <- which(df_balanced$Z == 1)
  
  ## group 0
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group0_first <- sample(idx_group0, 20)
  idx_group0 <- idx_group0[!idx_group0 %in% idx_group0_first]
  # the second batch to move
  idx_group0_second <- sample(idx_group0, floor(length(idx_group0)/2))
  # the third batch to move
  idx_group0_third <- idx_group0[!idx_group0 %in% idx_group0_second]
  
  # generate the dataframe for histogram animation
  # stage 3
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_first) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage3 <- hist_group0
  hist_group0_stage3$stage <- 3
  
  # stage 4
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_second) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage4 <- hist_group0
  hist_group0_stage4$stage <- 4
  
  
  # stage 5
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_third) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage5 <- hist_group0
  hist_group0_stage5$stage <- 5
  
  hist_group0_stage6 <- hist_group0_stage5
  hist_group0_stage6$stage <- 6
  hist_group0_stage7 <- hist_group0_stage5
  hist_group0_stage7$stage <- 7
  hist_group0_stage8 <- hist_group0_stage5
  hist_group0_stage8$stage <- 8
  
  hist_group0_stage1 <- hist_group0_stage5
  hist_group0_stage1$count <- 0
  hist_group0_stage1$stage <- 1
  
  hist_group0_stage2 <- hist_group0_stage5
  hist_group0_stage2$count <- 0
  hist_group0_stage2$stage <- 2
  
  histogram_group0 <- rbind(hist_group0_stage1, hist_group0_stage2, hist_group0_stage3, hist_group0_stage4, 
                            hist_group0_stage5, hist_group0_stage6, hist_group0_stage7, hist_group0_stage8)
  
  ## group 1
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group1_first <- sample(idx_group1, floor(length(idx_group1)/3))
  idx_group1 <- idx_group1[!idx_group1 %in% idx_group1_first]
  # the second batch to move
  idx_group1_second <- sample(idx_group1, floor(length(idx_group1)/2))
  # the third batch to move
  idx_group1_third <- idx_group1[!idx_group1 %in% idx_group1_second]
  
  # generate the dataframe for histogram animation
  # stage 6
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_first) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage6 <- hist_group1
  hist_group1_stage6$stage <- 6
  
  # stage 7
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_second) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage7 <- hist_group1
  hist_group1_stage7$stage <- 7
  
  # stage 8
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_third) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage8 <- hist_group1
  hist_group1_stage8$stage <- 8
  
  
  hist_group1_stage1 <- hist_group1_stage8
  hist_group1_stage1$count <- 0
  hist_group1_stage1$stage <- 1
  
  hist_group1_stage2 <- hist_group1_stage8
  hist_group1_stage2$count <- 0
  hist_group1_stage2$stage <- 2
  
  hist_group1_stage3 <- hist_group1_stage8
  hist_group1_stage3$count <- 0
  hist_group1_stage3$stage <- 3
  
  hist_group1_stage4 <- hist_group1_stage8
  hist_group1_stage4$count <- 0
  hist_group1_stage4$stage <- 4
  
  hist_group1_stage5 <- hist_group1_stage8
  hist_group1_stage5$count <- 0
  hist_group1_stage5$stage <- 5
  
  histogram_group1 <- rbind(hist_group1_stage1, hist_group1_stage2, hist_group1_stage3, hist_group1_stage4, 
                            hist_group1_stage5, hist_group1_stage6, hist_group1_stage7, hist_group1_stage8)
  
  
  # generate the dataframe for scatterplot animation
  df_balanced_stage1 <- df_balanced  
  df_balanced_stage1$stage <- 1
  df_balanced_stage1$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage2 <- df_balanced  
  df_balanced_stage2$Y[idx_group0_first] <- sapply(df_balanced_stage2$Y[idx_group0_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage2$stage <- 2
  df_balanced_stage2$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage3 <- df_balanced  
  df_balanced_stage3$Y[idx_group0_first] <- 0
  df_balanced_stage3$Y[idx_group0_second] <- sapply(df_balanced_stage2$Y[idx_group0_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage3$stage <- 3
  df_balanced_stage3$id <- 1:nrow(df_balanced)
  
  df_balanced_stage4 <- df_balanced  
  df_balanced_stage4$Y[idx_group0_first] <- 0
  df_balanced_stage4$Y[idx_group0_second] <- 0
  df_balanced_stage4$Y[idx_group0_third] <- sapply(df_balanced_stage2$Y[idx_group0_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage4$stage <- 4
  df_balanced_stage4$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage5 <- df_balanced  
  df_balanced_stage5$Y[idx_group0_first] <- 0
  df_balanced_stage5$Y[idx_group0_second] <- 0
  df_balanced_stage5$Y[idx_group0_third] <- 0
  df_balanced_stage5$Y[idx_group1_first] <- sapply(df_balanced_stage2$Y[idx_group1_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage5$stage <- 5
  df_balanced_stage5$id <- 1:nrow(df_balanced)
  
  df_balanced_stage6 <- df_balanced  
  df_balanced_stage6$Y[idx_group0_first] <- 0
  df_balanced_stage6$Y[idx_group0_second] <- 0
  df_balanced_stage6$Y[idx_group0_third] <- 0
  df_balanced_stage6$Y[idx_group1_first] <- 0
  df_balanced_stage6$Y[idx_group1_second] <- sapply(df_balanced_stage2$Y[idx_group1_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage6$stage <- 6
  df_balanced_stage6$id <- 1:nrow(df_balanced)
  
  df_balanced_stage7 <- df_balanced  
  df_balanced_stage7$Y[idx_group0_first] <- 0
  df_balanced_stage7$Y[idx_group0_second] <- 0
  df_balanced_stage7$Y[idx_group0_third] <- 0
  df_balanced_stage7$Y[idx_group1_first] <- 0
  df_balanced_stage7$Y[idx_group1_second] <- 0
  df_balanced_stage7$Y[idx_group1_third] <- sapply(df_balanced_stage2$Y[idx_group1_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage7$stage <- 7
  df_balanced_stage7$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage8 <- df_balanced  
  df_balanced_stage8$Y[idx_group0_first] <- 0
  df_balanced_stage8$Y[idx_group0_second] <- 0
  df_balanced_stage8$Y[idx_group0_third] <- 0
  df_balanced_stage8$Y[idx_group1_first] <- 0
  df_balanced_stage8$Y[idx_group1_second] <- 0
  df_balanced_stage8$Y[idx_group1_third] <- 0
  df_balanced_stage8$stage <- 8
  df_balanced_stage8$id <- 1:nrow(df_balanced)
  
  
  df <- rbind(df_balanced_stage1, df_balanced_stage2, df_balanced_stage3, df_balanced_stage4, 
              df_balanced_stage5, df_balanced_stage6, df_balanced_stage7, df_balanced_stage8)
  df$Z <- as.factor(df$Z)
  
  plot_ly() %>% 
    add_markers(data=df, x = ~X, y = ~Y, color = ~Z,colors = c("blue", "red"),  frame = ~stage, ids = ~id) %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group0, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'blue'
              ),  opacity = 0.4,name = '0') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group1, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'red'
              ),opacity = 0.4,name = '1') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_segments(x = min(df$X)-xaxis, xend = min(df$X)-xaxis, y = 0, yend = yaxis, 
                 size = I(1), line=list(color='black'), showlegend = F) %>% 
    layout(
      # title = 'Distribution of Observed Outcomes',
      font=list(
        family='Times New Roman',
        size=12),
      barmode = 'overlay', bargap = 0,
      xaxis = list(
        title = "",
        tick = list(
          size = 14,
          color = 'rgb(107, 107, 107)')))
}
```


Let's first look at an example where the covariate $X$ is imbalanced and does not complete overlap between the treatment and control groups.

```{r, echo=FALSE}
dgp_nonoverlap_unbalance <- function(n, seed){
  set.seed(seed)
  Z <- rbinom(n = n, size = 1, prob = 0.5)
  X <- ifelse(Z == 1, rnorm(n = n, mean = 40, sd = 10), rnorm(n = n, mean = 20, sd = 5))
  Y_0 <- 12 + 3*sqrt(X) + rnorm(n = n, 0,1)
  Y_1 <- 30 + exp(0.06*X) + rnorm(n = n, 0,1)
  Y <- ifelse(Z==1, Y_1, Y_0)
  output <- data.frame(Z, X, Y_0, Y_1, Y)
  colnames(output) <- c( "Z", "X", "Y0", "Y1", "Y")
  return(output)
}

df_nonoverlap_unbalance <- dgp_nonoverlap_unbalance(200, 123)
```

```{r, echo=FALSE, warning=FALSE}
animation(df_nonoverlap_unbalance, xaxis = 5, yaxis = 75)
```

```{r, echo=FALSE, message=FALSE}
p1 <- plot_marginal_distribution(df_nonoverlap_unbalance)
p2 <- plot_overlap(df_nonoverlap_unbalance)
ggarrange(p1, p2, nrow = 1, ncol = 2)
```

```{r, echo=FALSE}
SATE_meandiff <- paste('Mean Difference:', round(mean(df_nonoverlap_unbalance$Y[df_nonoverlap_unbalance$Z == 1]) - mean(df_nonoverlap_unbalance$Y[df_nonoverlap_unbalance$Z == 0]), 2))

SATE <- paste('True SATE:', round(mean(df_nonoverlap_unbalance$Y1 - df_nonoverlap_unbalance$Y0), 2))
```

```{r, echo=FALSE}
df_nonoverlap_unbalance$Z <- as.factor(df_nonoverlap_unbalance$Z)
model_nonoverlap_unbalance <- lm(Y ~ X + Z, data = df_nonoverlap_unbalance)
coef_nonoverlap_unbalance <- summary(model_nonoverlap_unbalance)$coefficients

inputPanel(selectInput("all_seeing_BART", label = "Choose to see researchers' or all-seeing perspectives:",
              choices = c("Researchers' view",
                          "All-seeing view"), 
              selected = "Researchers' view"))

renderPlot({
  colors <- c("0" = "blue", "1" = "red")
  linetypes <- c('Mean Difference' = 'solid', 'True SATE' = 'dashed', 'Linear Regression' = 'dotted')
  if(input$all_seeing_BART == "Researchers' view"){ 
    ggplot(data = df_nonoverlap_unbalance) + 
      # geom_line(aes(x = X, y = 30 + exp(0.06*X)), color = "red") +
      # geom_line(aes(x = X, y = 12 + 3*sqrt(X)), color = "blue") +
      geom_hline(aes(yintercept = mean(Y[Z == 1]), color = '1', linetype = 'Mean Difference')) +
      geom_hline(aes(yintercept = mean(Y[Z == 0]), color = '0', linetype = 'Mean Difference')) +
      geom_hline(aes(yintercept = mean(Y1), color = '1', linetype = 'True SATE')) +
      geom_hline(aes(yintercept = mean(Y0), color = '0', linetype = 'True SATE')) +
      geom_line(aes(x = X, y = coef_nonoverlap_unbalance[1] + coef_nonoverlap_unbalance[2] * X, linetype = 'Linear Regression')) + 
      geom_line(aes(x = X, y = coef_nonoverlap_unbalance[1] + coef_nonoverlap_unbalance[2] * X + coef_nonoverlap_unbalance[3],
                    linetype = 'Linear Regression')) +
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 1, ], aes(x = X, y = Y, color = "1")) + 
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 0, ], aes(x = X, y = Y, color = "0")) + 
      scale_color_manual("Treated?", values = colors) + 
      scale_linetype_manual("Linetype", values = linetypes) +
      theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") + ylab("Observed Outcome") +
      annotate(
        "text", label = SATE_meandiff,
        x = 21, y = 65, size = 4, colour = "red"
      ) + 
      annotate(
        "text", label = SATE,
        x = 20, y = 62, size = 4, colour = "red"
      ) + 
      guides(color = guide_legend(order = 1), linetype = guide_legend(order = 2))
  }else{
    shapes <- c('Observed' = 16, 'Counterfactual' = 1)
    ggplot(data = df_nonoverlap_unbalance) + 
      geom_line(aes(x = X, y = 30 + exp(0.06*X)), color = "red") +
      geom_line(aes(x = X, y = 12 + 3*sqrt(X)), color = "blue") +
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 1, ], aes(x = X, y = Y, color = "1", shape = 'Observed')) + 
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 0, ], aes(x = X, y = Y, color = "0", shape = 'Observed')) + 
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 1, ], aes(x = X, y = Y0, color = "1", shape = 'Counterfactual')) +
      geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 0, ], aes(x = X, y = Y1, color = "0", shape = 'Counterfactual')) +
      scale_color_manual("Treated?", values = colors) + scale_shape_manual('Shape', values = shapes) +
      theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") + ylab("Observed Outcome") +
      guides(color = guide_legend(order = 1), shape = guide_legend(order = 2))
    
  }
})
```

The difference in means is a biased estimate of the treatment effect. The OLS estimator also fails to capture the true effect because there just isn't enough information in the data to extrapolate fully into areas where there isn't overlap. Using flexible models such as BART would be the way to go. The BART point estimate (posterior mean) of the average effect of the treatment is 8.07 with 95% posterior interval (6.14, 9.99), much closer to the true SATE, 10.01.

```{r, echo=FALSE}
# linear regression
model_nonoverlap_unbalance <- lm(Y ~ X + Z, data = df_nonoverlap_unbalance)
coef_nonoverlap_unbalance <- summary(model_nonoverlap_unbalance)$coefficients
  
inputPanel(selectInput("bart_select_lm_result", h4("Select the result from the linear regression to show: "),
                       choices = c("Only show the coefficient estimated for the treatment effect" = "bart_lm_coefficient",
                                   "Show the full regression output" = "bart_lm_full_output"),
                       selected = "bart_lm_coefficient"),
           conditionalPanel("input.bart_select_lm_result === 'bart_lm_full_output'",
                            verbatimTextOutput('bart_lm_result')),
           conditionalPanel("input.bart_select_lm_result === 'bart_lm_coefficient'",
                            verbatimTextOutput('bart_lm_coef')))

output$bart_lm_result <- renderPrint({
  summary(model_nonoverlap_unbalance)
})
output$bart_lm_coef <- renderPrint({
  round(coef_nonoverlap_unbalance[,1]["Z1"],2)
})
```

```{r, message=FALSE, echo=FALSE}
# BART
set.seed(123) # doesn't work,  result below different every time
df_nonoverlap_unbalance$Z_val <- as.numeric(as.character(df_nonoverlap_unbalance$Z))
fit <- bartc(response = Y, treatment = Z_val, confounders = X, data = df_nonoverlap_unbalance)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
inputPanel(selectInput("bart_result", h4("Select the result from the BART model to show: "),
                       choices = c("Only show the coefficient estimated for the treatment effect" = "bart_coefficient",
                                   "Show the full regression output" = "bart_full_output"),
                       selected = "bart_coefficient"),
           conditionalPanel("input.bart_result === 'bart_full_output'",
                            verbatimTextOutput('bart_result')),
           conditionalPanel("input.bart_result === 'bart_coefficient'",
                            verbatimTextOutput('bart_coef')))

output$bart_result <- renderPrint({
  summary(fit)
})
output$bart_coef <- renderPrint({
  round(summary(fit)$estimates,2)
})
```


The dashed line in the figure below displays the BART fit to the data which is quite close to the true conditional expectation for most of the support except at values of X far from the area of strong overlap.

```{r, echo=FALSE}
y0_hat <- fitted(object = fit, type = 'y.0')
y1_hat <- fitted(object = fit, type = 'y.1')

df_BART <- data.frame(x=df_nonoverlap_unbalance$X, y0 = y0_hat, y1 = y1_hat)
colors <- c("0" = "blue", "1" = "red")
ggplot() + geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 1, ], aes(x = X, y = Y, color = "1")) + 
  geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 0, ], aes(x = X, y = Y, color = "0")) + 
  geom_line(data = df_nonoverlap_unbalance, aes(x = X, y = 30 + exp(0.06*X)), color = "red") +
  geom_line(data = df_nonoverlap_unbalance, aes(x = X, y = 12 + 3*sqrt(X)), color = "blue") +
  scale_color_manual("Treated?", values=colors) + 
  theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") + 
  geom_line(data = df_BART, aes(x = x, y = y0), linetype = 'dashed') +
  geom_line(data = df_BART, aes(x = x, y = y1), linetype = 'dashed') 
```

The extrapolations of BART over areas of the covariate space where common support does not exist can lead to biased inferences because of the lack of information available to identify either E[Y(0) | X] or E[Y(1) | X] in these regions. After discarding observations in neighborhoods of the covariate space that lack sufficient common causal support using the BART rule 'sd', the interval cover the true treatment effect for the remaining sample.

```{r, message=FALSE, echo=FALSE, warning=FALSE}
fit1 <- bartc(response = Y, treatment = Z_val, confounders = X, data = df_nonoverlap_unbalance, commonSup.rule = 'sd')
inputPanel(selectInput("bart_discard_result", h4("Select the result from the BART model to show: "),
                       choices = c("Only show the coefficient estimated for the treatment effect" = "bart_discard_coefficient",
                                   "Show the full regression output" = "bart_discard_full_output"),
                       selected = "bart_discard_coefficient"),
           conditionalPanel("input.bart_discard_result === 'bart_discard_full_output'",
                            verbatimTextOutput('bart_discard_result')),
           conditionalPanel("input.bart_discard_result === 'bart_discard_coefficient'",
                            verbatimTextOutput('bart_discard_coef')))

output$bart_discard_result <- renderPrint({
  summary(fit1)
})
output$bart_discard_coef <- renderPrint({
  round(summary(fit1)$estimates,2)
})
```

```{r, echo=FALSE}
df_commonSup <- df_nonoverlap_unbalance[fit1$commonSup.sub,]
SATT <- mean(df_commonSup[df_commonSup$Z==1,]$Y1 - df_commonSup[df_commonSup$Z==1,]$Y0)
cat('True SATT = ', SATT, '\n')
```

```{r, echo=FALSE}
model_1 <- lm(Y1~X, data = df_commonSup)
model_0 <- lm(Y0~X, data = df_commonSup)
CATT <- mean(model_1$fitted.values[df_commonSup$Z==1] - model_0$fitted.values[df_commonSup$Z==1])
cat('True CATT = ', CATT)
```

SATT and CATT for the remaining units are 7.87 and 7.95, respectively. Our new BART estimate is 7.62 with 95% posterior interval (6.9, 8.3), which covers the SATT and CATT. In the left panel of the figure below, points in circle are the observations discarded based on the BART rule. The gray vertical segments are 95% intervals for the posterior predicted counterfactuals at each $X$ value from an observation. The right panel displays the BART inference for each treated unit (which can be averaged to estimate the effect of the treatment on the treated for this sample). In the right panel, the true treatment effect as it varies with $X$, $E[Y(1)−Y(0) | X]$, is plotted as the solid curve. The vertical segments are marginal 95% posterior intervals for the treatment effect at each $X$ value from a treated observation. Notice that the uncertainty bounds grow much wider in the range where there is no overlap across treatment groups.


```{r, echo=FALSE}
y0_hat <- fitted(object = fit1, type = 'y.0')
y1_hat <- fitted(object = fit1, type = 'y.1')
y.cf <- extract(fit1, 'y.cf')
y.point <- apply(y.cf, 2, mean)
y.sd <- apply(y.cf, 2, sd)
ucl <- y.point + 2*y.sd 
lcl<- y.point - 2*y.sd

df_BART <- data.frame(x=df_nonoverlap_unbalance$X, y = df_nonoverlap_unbalance$Y, y0 = y0_hat, y1 = y1_hat, y.point = y.point, y.sd = y.sd, ucl = ucl, lcl = lcl)
```


```{r, echo=FALSE}
p1 <- ggplot() + geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 1, ], aes(x = X, y = Y, color = "1")) + 
  geom_point(data = df_nonoverlap_unbalance[df_nonoverlap_unbalance$Z == 0, ], aes(x = X, y = Y, color = "0")) + 
  geom_line(data = df_nonoverlap_unbalance, aes(x = X, y = 30 + exp(0.06*X)), color = "red") +
  geom_line(data = df_nonoverlap_unbalance, aes(x = X, y = 12 + 3*sqrt(X)), color = "blue") +
  scale_color_manual("Treated?", values=colors) + 
  theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") + 
  geom_line(data = df_BART, aes(x = x, y = y0), linetype = 'dashed') +
  geom_line(data = df_BART, aes(x = x, y = y1), linetype = 'dashed') +
  geom_circle(data = df_nonoverlap_unbalance[!fit1$commonSup.sub,], aes(x0 = X, y0 = Y, r = 0.6)) + 
  geom_errorbar(data = df_BART, aes(x = x, y = y, ymin = lcl, ymax = ucl), color = 'lightgrey')
```


```{r, echo=FALSE}
ites <- extract(fit1, type = "ite")
ite.m <- apply(ites, 2, mean)
ite.sd <- apply(ites, 2, sd)
ite.lb <- ite.m - 2 * ite.sd
ite.ub <- ite.m + 2 * ite.sd

df_ite <- data.frame(x=df_nonoverlap_unbalance$X, ucl = ite.lb, lcl = ite.ub)
idx <- which(df_nonoverlap_unbalance$Z == 1)
p2 <- ggplot() + 
  geom_line(data = df_nonoverlap_unbalance[idx, ], 
            aes(x = X, y = 30 + exp(0.06*X) - 12 - 3*sqrt(X), color = "red")) +
  geom_errorbar(data = df_ite[idx,], aes(x = x, ymin = lcl, ymax = ucl), color = 'grey') +
  theme_minimal(base_line_size = 0.4) + ylab('treatment effect') + 
  theme(legend.position="bottom", legend.title = element_text(color = "transparent"), 
        legend.text = element_text(color = "transparent")) + 
    scale_color_discrete(guide = guide_legend(override.aes = list(alpha = 0)))

ggarrange(p1, p2, nrow = 1, ncol = 2)
```



