---
title: "Overlap"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(ggpubr)
library(tidyverse)
library(shiny)
library(bartCause)
library(ggforce)
library(gridExtra)
library(cowplot)
library(ggthemes) 
library(plotly)
```

```{r, echo=FALSE}
plot_scatterplot <- function(df){
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  linetypes <- c('Mean Difference' = 'solid', 'True SATE' = 'dashed')
  ggplot(df) + 
    geom_point( aes(x = X, y = Y, color = Z ))+ 
    xlab("X") + ylab("Observed Outcome") + 
    labs(title = "Distribution of Observed Outcomes") + 
    geom_hline(aes(yintercept = mean(Y[Z == 1]), color = '1', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y[Z == 0]), color = '0', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y1), color = '1', linetype = 'True SATE')) +
    geom_hline(aes(yintercept = mean(Y0), color = '0', linetype = 'True SATE')) +
    scale_color_manual("Treated?", values = colors) + 
    scale_linetype_manual("Linetype", values = linetypes) +
    theme_minimal(base_line_size = 0.4) + 
    theme(plot.title = element_text(size = 8), 
          legend.position="bottom",
          text=element_text(family="serif", size=8)) + 
    guides(color = guide_legend(order = 1), fill = guide_legend(order = 2))
    
}

plot_marginal_distribution <- function(df) {
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  ggplot(df, aes(x = X, fill = Z)) + 
    theme_minimal(base_line_size = 0.4) +
    geom_histogram(bins = 60, alpha = 0.4, position = "identity") +
    geom_vline(data = df[which(df$Z == 0),], aes(xintercept= mean(X), color="0"), size=1) + 
    geom_vline(data = df[which(df$Z == 1),], aes(xintercept= mean(X), color="1"), size=1) +
    scale_color_manual(name = "Treated?", values = colors) + 
    scale_fill_manual(name = "Treated?", values = colors) +
    labs(title = "Distribution of Covariate X", y = "Frequency") +
    theme(legend.position="bottom",
          plot.title = element_text(size = 8), 
          text=element_text(family="serif", size=8)) 
    
}

plot_overlap <- function(df, nooverlap = FALSE){
  p <- ggplot(data = df) + geom_histogram(bins = 60, aes(x = X, color = Z, fill = Z), alpha = 0.5)
  get_hist <- function(p) {
    d <- ggplot_build(p)$data[[1]]
    data.frame(x = d$x, xmin = d$xmin, xmax = d$xmax, y = d$y, mids = (d$xmin + d$xmax)/2)
  }
  hist = get_hist(p)
  df$Z <- as.factor(df$Z)
  if(nooverlap == TRUE){
    ggplot() + 
      geom_histogram(data = df %>% filter(Z == 1),   
                     aes( x = X, y = ..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) +
      geom_histogram(data = df %>% filter(Z == 0), 
                     aes(x = X, y = -..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) + 
      labs(title = "Distribution of Covariate X", y = "Frequency") + 
      scale_fill_manual("Treated?", values=c("blue","red")) + theme_minimal(base_line_size = 0.4) +
      theme(legend.position="bottom", plot.title = element_text(size=8),
            text=element_text(family="serif", size=8)) 
  }else{
    ggplot() + 
      geom_histogram(data = df %>% filter(Z == 1), 
                     aes( x = X, y = ..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) +
      geom_histogram(data = df %>% filter(Z == 0), 
                     aes(x = X, y = -..count.., fill = Z), alpha = 0.8, binwidth = (hist$xmax-hist$xmin)[1]) + 
      labs(title = "Distribution of Covariate X", y = "Frequency") + 
      scale_fill_manual("Treated?", values=c("blue","red")) + theme_minimal(base_line_size = 0.4) +
      theme(legend.position="bottom", plot.title = element_text(size=8),
          text=element_text(family="serif", size=8)) 
  }
}

dgp_radomized_balanced <- function (n, beta_1, seed) {
  set.seed(seed) 
  pre_quiz <- rnorm(n = n, mean = 65, sd = 3)
  beta_0 <- 10
  tau <- 5
  Z <- sample(x = c(0,1), size = n, replace = TRUE)
  Y_0 <- beta_0 + beta_1 * pre_quiz + 0 + rnorm(n) 
  Y_1 <- beta_0 + beta_1 * pre_quiz + tau + rnorm(n) 
  Y <- ifelse(Z == 1, Y_1, Y_0)
  output <- data.frame(Z, pre_quiz, Y_0, Y_1, Y)
  colnames(output) <- c( "Z", "X", "Y0", "Y1", "Y")
  return (output) 
}
```

```{r, echo=FALSE}
animation <- function(df_balanced, xaxis, yaxis){
  p <- ggplot(data = df_balanced) + geom_histogram(bins = 60, aes(x = X, color = Z, fill = Z), alpha = 0.5)
  get_hist <- function(p) {
    d <- ggplot_build(p)$data[[1]]
    data.frame(x = d$x, xmin = d$xmin, xmax = d$xmax, y = d$y, mids = (d$xmin + d$xmax)/2)
  }
  hist = get_hist(p)
  hist_group0 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  hist_group1 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  
  idx_group0 <- which(df_balanced$Z == 0)
  idx_group1 <- which(df_balanced$Z == 1)
  
  ## group 0
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group0_first <- sample(idx_group0, 20)
  idx_group0 <- idx_group0[!idx_group0 %in% idx_group0_first]
  # the second batch to move
  idx_group0_second <- sample(idx_group0, floor(length(idx_group0)/2))
  # the third batch to move
  idx_group0_third <- idx_group0[!idx_group0 %in% idx_group0_second]
  
  # generate the dataframe for histogram animation
  # stage 3
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_first) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage3 <- hist_group0
  hist_group0_stage3$stage <- 3
  
  # stage 4
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_second) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage4 <- hist_group0
  hist_group0_stage4$stage <- 4
  
  
  # stage 5
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_third) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage5 <- hist_group0
  hist_group0_stage5$stage <- 5
  
  hist_group0_stage6 <- hist_group0_stage5
  hist_group0_stage6$stage <- 6
  hist_group0_stage7 <- hist_group0_stage5
  hist_group0_stage7$stage <- 7
  hist_group0_stage8 <- hist_group0_stage5
  hist_group0_stage8$stage <- 8
  
  hist_group0_stage1 <- hist_group0_stage5
  hist_group0_stage1$count <- 0
  hist_group0_stage1$stage <- 1
  
  hist_group0_stage2 <- hist_group0_stage5
  hist_group0_stage2$count <- 0
  hist_group0_stage2$stage <- 2
  
  histogram_group0 <- rbind(hist_group0_stage1, hist_group0_stage2, hist_group0_stage3, hist_group0_stage4, 
                            hist_group0_stage5, hist_group0_stage6, hist_group0_stage7, hist_group0_stage8)
  
  ## group 1
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group1_first <- sample(idx_group1, floor(length(idx_group1)/3))
  idx_group1 <- idx_group1[!idx_group1 %in% idx_group1_first]
  # the second batch to move
  idx_group1_second <- sample(idx_group1, floor(length(idx_group1)/2))
  # the third batch to move
  idx_group1_third <- idx_group1[!idx_group1 %in% idx_group1_second]
  
  # generate the dataframe for histogram animation
  # stage 6
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_first) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage6 <- hist_group1
  hist_group1_stage6$stage <- 6
  
  # stage 7
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_second) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage7 <- hist_group1
  hist_group1_stage7$stage <- 7
  
  # stage 8
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_third) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage8 <- hist_group1
  hist_group1_stage8$stage <- 8
  
  
  hist_group1_stage1 <- hist_group1_stage8
  hist_group1_stage1$count <- 0
  hist_group1_stage1$stage <- 1
  
  hist_group1_stage2 <- hist_group1_stage8
  hist_group1_stage2$count <- 0
  hist_group1_stage2$stage <- 2
  
  hist_group1_stage3 <- hist_group1_stage8
  hist_group1_stage3$count <- 0
  hist_group1_stage3$stage <- 3
  
  hist_group1_stage4 <- hist_group1_stage8
  hist_group1_stage4$count <- 0
  hist_group1_stage4$stage <- 4
  
  hist_group1_stage5 <- hist_group1_stage8
  hist_group1_stage5$count <- 0
  hist_group1_stage5$stage <- 5
  
  histogram_group1 <- rbind(hist_group1_stage1, hist_group1_stage2, hist_group1_stage3, hist_group1_stage4, 
                            hist_group1_stage5, hist_group1_stage6, hist_group1_stage7, hist_group1_stage8)
  
  
  # generate the dataframe for scatterplot animation
  df_balanced_stage1 <- df_balanced  
  df_balanced_stage1$stage <- 1
  df_balanced_stage1$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage2 <- df_balanced  
  df_balanced_stage2$Y[idx_group0_first] <- sapply(df_balanced_stage2$Y[idx_group0_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage2$stage <- 2
  df_balanced_stage2$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage3 <- df_balanced  
  df_balanced_stage3$Y[idx_group0_first] <- 0
  df_balanced_stage3$Y[idx_group0_second] <- sapply(df_balanced_stage2$Y[idx_group0_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage3$stage <- 3
  df_balanced_stage3$id <- 1:nrow(df_balanced)
  
  df_balanced_stage4 <- df_balanced  
  df_balanced_stage4$Y[idx_group0_first] <- 0
  df_balanced_stage4$Y[idx_group0_second] <- 0
  df_balanced_stage4$Y[idx_group0_third] <- sapply(df_balanced_stage2$Y[idx_group0_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage4$stage <- 4
  df_balanced_stage4$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage5 <- df_balanced  
  df_balanced_stage5$Y[idx_group0_first] <- 0
  df_balanced_stage5$Y[idx_group0_second] <- 0
  df_balanced_stage5$Y[idx_group0_third] <- 0
  df_balanced_stage5$Y[idx_group1_first] <- sapply(df_balanced_stage2$Y[idx_group1_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage5$stage <- 5
  df_balanced_stage5$id <- 1:nrow(df_balanced)
  
  df_balanced_stage6 <- df_balanced  
  df_balanced_stage6$Y[idx_group0_first] <- 0
  df_balanced_stage6$Y[idx_group0_second] <- 0
  df_balanced_stage6$Y[idx_group0_third] <- 0
  df_balanced_stage6$Y[idx_group1_first] <- 0
  df_balanced_stage6$Y[idx_group1_second] <- sapply(df_balanced_stage2$Y[idx_group1_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage6$stage <- 6
  df_balanced_stage6$id <- 1:nrow(df_balanced)
  
  df_balanced_stage7 <- df_balanced  
  df_balanced_stage7$Y[idx_group0_first] <- 0
  df_balanced_stage7$Y[idx_group0_second] <- 0
  df_balanced_stage7$Y[idx_group0_third] <- 0
  df_balanced_stage7$Y[idx_group1_first] <- 0
  df_balanced_stage7$Y[idx_group1_second] <- 0
  df_balanced_stage7$Y[idx_group1_third] <- sapply(df_balanced_stage2$Y[idx_group1_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage7$stage <- 7
  df_balanced_stage7$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage8 <- df_balanced  
  df_balanced_stage8$Y[idx_group0_first] <- 0
  df_balanced_stage8$Y[idx_group0_second] <- 0
  df_balanced_stage8$Y[idx_group0_third] <- 0
  df_balanced_stage8$Y[idx_group1_first] <- 0
  df_balanced_stage8$Y[idx_group1_second] <- 0
  df_balanced_stage8$Y[idx_group1_third] <- 0
  df_balanced_stage8$stage <- 8
  df_balanced_stage8$id <- 1:nrow(df_balanced)
  
  
  df <- rbind(df_balanced_stage1, df_balanced_stage2, df_balanced_stage3, df_balanced_stage4, 
              df_balanced_stage5, df_balanced_stage6, df_balanced_stage7, df_balanced_stage8)
  df$Z <- as.factor(df$Z)
  
  plot_ly() %>% 
    add_markers(data=df, x = ~X, y = ~Y, color = ~Z,colors = c("blue", "red"),  frame = ~stage, ids = ~id) %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group0, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'blue'
              ),  opacity = 0.4,name = '0') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group1, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'red'
              ),opacity = 0.4,name = '1') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_segments(x = min(df$X)-xaxis, xend = min(df$X)-xaxis, y = 0, yend = yaxis, 
                 size = I(1), line=list(color='black'), showlegend = F) %>% 
    layout(
      # title = 'Distribution of Observed Outcomes',
      font=list(
        family='Times New Roman',
        size=12),
      barmode = 'overlay', bargap = 0,
      xaxis = list(
        title = "",
        tick = list(
          size = 14,
          color = 'rgb(107, 107, 107)')))
}
```


Overlap or common support describes the extent to which the support of the covariate data is the same between the treatment and control groups. There is complete overlap when there exist both treatment and control units in all neighborhoods of the covariate space. 

```{r, echo=FALSE}
df_balanced <- dgp_radomized_balanced(n = 1000, beta_1 = 1.1, seed = 14)
```

```{r, warning=FALSE, echo=FALSE}
animation(df_balanced, xaxis = 1.5, yaxis = 100)
```

```{r, echo=FALSE}
SATE_meandiff <- paste('Mean Difference:', round(mean(df_balanced[df_balanced$Z == 1, ]$Y) - mean(df_balanced[df_balanced$Z == 0, ]$Y), 2))
SATE <- paste('True SATE:', round(mean(df_balanced$Y1 - df_balanced$Y0), 2))
```

```{r, echo=FALSE}
p1 <- plot_scatterplot(df_balanced) +
  annotate(
    "text", label = SATE_meandiff,
    x = 60, y = 95, size = 3, colour = "red"
  ) + 
  annotate(
    "text", label = SATE,
    x = 60, y = 94, size = 3, colour = "red"
  )
p1_overlap <- plot_overlap(df_balanced, nooverlap = T)
plot_grid(p1,  p1_overlap, ncol = 2, nrow = 1)
```

Lack of complete overlap in confounders creates problems, because in that setting there are treatment observations for which we have no empirical counterfactuals (that is, control observations with the same covariate distribution) or vice versa. (a) Two distributions with no overlap; (b) two distributions with partial overlap; (c) a scenario in which the range of one distribution is a subset of the range of the other

```{r, echo=FALSE}
dgp_overlap <- function(n, delta_c = 5, a1 = 0.5, b1 = 5, a0 = -0.5, b0 = 5, 
                        lwr, upp, seed = 100, nooverlap = FALSE, subset_flag = FALSE){
  set.seed(seed)
  X <- rnorm(n = n, mean = 0, sd = 3)
  # Compare the log-odds scale of lower and upper bounds and return the lowest
  lwr_pre <- log(lwr / (1 - lwr))
  upp_pre <- log(upp / (1 - upp))
  lwr_alpha <- lwr_pre / min(X)
  upp_alpha <- upp_pre / max(X)
  alpha <- min(lwr_alpha, upp_alpha)
  p <- exp(alpha * X) / (1 + exp(alpha * X))
  if(nooverlap){
    Z <- ifelse(p>=0.5, 1, 0)
  }else if (subset_flag){
    Z <- c(rbinom(n = floor(n * 0.6), size = 1, prob = p[1:floor(n * 0.6)]), rep(0, n - floor(n * 0.6)))
  }else{
    Z <- rbinom(n = n, size = 1, prob = p)
  }
  Y0 <- 60 + a0 * X^2 + b0 * X + rnorm(n = n, mean = 0, sd = 2)
  Y1 <- 60 + a1 * X^2 + b1 * X + delta_c + rnorm(n = n, mean = 0, sd = 2)
  Y <- ifelse( Z == 1, Y1, Y0)
  return(data.frame( X, Y, Y0, Y1, Z))
}

delta_c = 5
a1 = 0.5
b1 = 5
a0 = -0.5
b0 = 5

df_overlap3 <- dgp_overlap(n = 100, delta_c = delta_c, a1 = a1, b1 = b1, a0 = a0, b0 = b0, 
                           lwr = 0.05, upp = 0.95, nooverlap = TRUE)
df_overlap2 <- dgp_overlap(n = 100, delta_c = delta_c, a1 = a1, b1 = b1, a0 = a0, b0 = b0, 
                           lwr = 0.05, upp = 0.95)
df_overlap1 <- dgp_overlap(n = 100, delta_c = delta_c, a1 = a1, b1 = b1, a0 = a0, b0 = b0, 
                           lwr = 0.05, upp = 0.95, subset_flag = TRUE)
```


```{r, warning=FALSE, echo=FALSE}
animation(df_overlap3, xaxis = 1.5, yaxis = 142)
animation(df_overlap2, xaxis = 1.5, yaxis = 125)
animation(df_overlap1, xaxis = 1.5, yaxis = 125)
```

```{r, echo=FALSE}
SATE_meandiff_left <- paste('Mean Difference:', round(mean(df_overlap3$Y[df_overlap3$Z == 1]) - mean(df_overlap3$Y[df_overlap3$Z == 0]), 2))
SATE_left <- paste('True SATE:', round(mean(df_overlap3$Y1 - df_overlap3$Y0), 2))

SATE_meandiff_middle <- paste('Mean Difference:', round(mean(df_overlap2$Y[df_overlap2$Z == 1]) - mean(df_overlap2$Y[df_overlap2$Z == 0]), 2))
SATE_middle <- paste('True SATE:', round(mean(df_overlap2$Y1 - df_overlap2$Y0), 2))

SATE_meandiff_right <- paste('Mean Difference:', round(mean(df_overlap1$Y[df_overlap1$Z == 1]) - mean(df_overlap1$Y[df_overlap1$Z == 0]), 2))
SATE_right <- paste('True SATE:', round(mean(df_overlap1$Y1 - df_overlap1$Y0), 2))
```

```{r, echo=FALSE}
p1 <- plot_scatterplot(df_overlap1) + theme(plot.title = element_text(size=8)) + 
  annotate(
    "text", label = SATE_meandiff_right,
    x = -3, y = 120, size = 2, colour = "red"
  ) + 
  annotate(
    "text", label = SATE_right,
    x = -3.6, y = 115, size = 2, colour = "red"
  )
p2 <- plot_scatterplot(df_overlap2) + theme(plot.title = element_text(size=8)) + 
  annotate(
    "text", label = SATE_meandiff_middle,
    x = -3, y = 125, size = 2, colour = "red"
  ) + 
  annotate(
    "text", label = SATE_middle,
    x = -3.6, y = 120, size = 2, colour = "red"
  )
p3 <- plot_scatterplot(df_overlap3) + theme(plot.title = element_text(size=8)) + 
  annotate(
    "text", label = SATE_meandiff_left,
    x = -3, y = 125, size = 2, colour = "red"
  ) + 
  annotate(
    "text", label = SATE_left,
    x = -3.6, y = 120, size = 2, colour = "red"
  )
ggarrange(p3, p2, p1, nrow = 1, ncol = 3)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p1 <- plot_overlap(df_overlap1) + theme(plot.title = element_text(size=8))
p2 <- plot_overlap(df_overlap2) + theme(plot.title = element_text(size=8))
p3 <- plot_overlap(df_overlap3, nooverlap = T) + theme(plot.title = element_text(size=8))
ggarrange(p3, p2, p1, nrow = 1, ncol = 3)
```

Suppose we are interested in estimating the treatment effect in the middle example above, and in the example $X$ is only one confounding covariate — that is, only one predictor $X$ is necessary to satisfy ignorability. The red dots correspond to the units who received the treatment; the blue dots correspond to the units who did not receive the treatment. The dashed lines are regression lines fit to the observed data. Because there are no control units with high X and no treatment units with low $X$, the linear model, to create counterfactual predictions, is forced to extrapolate over portions of the space where there are no data to support them. This causes an underestimate of the true average treatment effect. Allowing for an interaction does not solve the problem. You may see in the all-seeing plot where the red solid line represents the true relationship between the potential outcome for treatment receipt and the $X$, $E(Y1|X)$, and the blue solid line represents the true relationship between the potential outcome for the control condition and the $X$, $E(Y0|X)$, the true causal effect at any level of the X is the vertical distance between the two solid lines. Each average causal effect is an average across the relevant subset of these individual-level causal effects. 

```{r, echo=FALSE}
inputPanel(selectInput("all_seeing", label = "Choose to see researchers' or all-seeing perspectives:",
              choices = c("Researchers' view - fitting a linear regression", 
                          "Researchers' view - fitting a linear regression with an interaction", 
                          "All-seeing view"), 
              selected = "Researchers' view - fitting a linear regression"))


df_overlap2$Z <- as.factor(df_overlap2$Z)
model <- lm(Y ~ X + Z, data = df_overlap2)
model_interaction <- lm(Y ~ X*Z, data = df_overlap2)
coef_model <- summary(model)$coefficients
coef_model_interaction <- summary(model_interaction)$coefficients
colors <- c("0" = "blue", "1" = "red")
linetype <- c("linear model without interaction" = "dashed", "linear model with an interaction" = "dotted")

renderPlot({
   if(input$all_seeing == "Researchers' view - fitting a linear regression"){ 
     ggplot() + geom_point(data = df_overlap2[df_overlap2$Z == 1, ], aes(x = X, y = Y, color = "1")) + 
      geom_point(data = df_overlap2[df_overlap2$Z == 0, ], aes(x = X, y = Y, color = "0")) + 
      scale_color_manual("Treated?", values=colors) + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model[1] + coef_model[2] * X), linetype = "dashed") + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model[1] + coef_model[2] * X + coef_model[3]), linetype = "dashed") +
      theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") 
   }else if(input$all_seeing == "Researchers' view - fitting a linear regression with an interaction"){
     ggplot() + geom_point(data = df_overlap2[df_overlap2$Z == 1, ], aes(x = X, y = Y, color = "1")) + 
      geom_point(data = df_overlap2[df_overlap2$Z == 0, ], aes(x = X, y = Y, color = "0")) + 
      scale_color_manual("Treated?", values=colors) + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model_interaction[1] + coef_model_interaction[2] * X), 
                linetype = "dotted") + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model_interaction[1] + 
               (coef_model_interaction[2] + coef_model_interaction[4]) * X + coef_model_interaction[3]), 
               linetype = "dotted") + theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom")
   }else{ # all-seeing view
     shapes <- c('Observed' = 16, 'Counterfactual' = 1)
     ggplot() + geom_point(data = df_overlap2[df_overlap2$Z == 1, ], aes(x = X, y = Y, color = "1", shape = 'Observed')) + 
      geom_point(data = df_overlap2[df_overlap2$Z == 0, ], aes(x = X, y = Y, color = "0", shape = 'Observed')) + 
      geom_point(data = df_overlap2[df_overlap2$Z == 1, ], aes(x = X, y = Y0, color = "1", shape = 'Counterfactual')) +
      geom_point(data = df_overlap2[df_overlap2$Z == 0, ], aes(x = X, y = Y1, color = "0", shape = 'Counterfactual')) +
      geom_line(data = df_overlap2, aes(x = X, y = 60 + b1*X + a1*X^2 + delta_c), color = "red") +
      geom_line(data = df_overlap2, aes(x = X, y = 60 + b0*X + a0*X^2), color = "blue") +
      scale_color_manual("Treated?", values=colors) + 
      scale_shape_manual('Shape', values = shapes) +
      geom_line(data = df_overlap2, aes(x = X, y = coef_model[1] + coef_model[2] * X, linetype = "linear model without interaction")) + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model[1] + coef_model[2] * X + coef_model[3], 
                                        linetype = "linear model without interaction")) +
      geom_line(data = df_overlap2, aes(x = X, y = coef_model_interaction[1] + coef_model_interaction[2] * X, 
                                        linetype = "linear model with an interaction")) + 
      geom_line(data = df_overlap2, aes(x = X, y = coef_model_interaction[1] + (coef_model_interaction[2] + coef_model_interaction[4]) * X +
                                          coef_model_interaction[3], linetype = "linear model with an interaction")) +
       theme_minimal(base_line_size = 0.4) + theme(legend.position="bottom") + scale_linetype_manual( "Linetype", values=linetype)
   }
 })

```


```{r, echo=FALSE}
inputPanel( selectInput("nonoverlap_select_lm_result", h4("Select the result from the linear regression to show: "),
                       choices = c("Only show the coefficient estimated for the treatment effect" = "nonoverlap_lm_coefficient",
                                   "Show the full regression output" = "nonoverlap_lm_full_output"),
                       selected = "nonoverlap_lm_coefficient"),
           conditionalPanel("input.nonoverlap_select_lm_result === 'nonoverlap_lm_full_output'",
                            verbatimTextOutput('nonoverlap_lm_result')),
           conditionalPanel("input.nonoverlap_select_lm_result === 'nonoverlap_lm_coefficient'",
                            verbatimTextOutput('nonoverlap_lm_coef')))

output$nonoverlap_lm_result <- renderPrint({
  summary(model)
})
output$nonoverlap_lm_coef <- renderPrint({
  round(coef_model[,1]["Z1"],2)
})
```


```{r, echo=FALSE}
inputPanel(selectInput("nonoverlap_select_lm_interaction_result", h4("Select the result from the linear regression with interaction to show: "),
                       choices = c("Only show the coefficient estimated for the treatment effect" = "nonoverlap_lm_interaction_coefficient",
                                   "Show the full regression output" = "nonoverlap_lm_interaction_full_output"),
                       selected = "nonoverlap_lm_interaction_coefficient"),
           conditionalPanel("input.nonoverlap_select_lm_interaction_result === 'nonoverlap_lm_interaction_full_output'",
                            verbatimTextOutput('nonoverlap_lm_interaction_result')),
           conditionalPanel("input.nonoverlap_select_lm_interaction_result === 'nonoverlap_lm_interaction_coefficient'",
                            verbatimTextOutput('nonoverlap_lm_interaction_coef')))

output$nonoverlap_lm_interaction_result <- renderPrint({
  summary(model_interaction)
})
output$nonoverlap_lm_interaction_coef <- renderPrint({
  round(coef_model_interaction[,1]["Z1"],2)
})
```

```{r, echo=FALSE}
SATE <- mean(df_overlap2$Y1 - df_overlap2$Y0)
cat('True SATE:', SATE)
```

Since we rely on empirical counterfactual units to inform counterfactual outcomes, when treatment and control groups do not completely overlap, the data are inherently limited in what they can tell us about treatment effects in the regions of nonoverlap. In regions where the groups do not have covariate overlap, causal estimation is purely based on extrapolation. Hence, any inferences in the areas with no overlap would be sensitive to model-specification. Failure to detect areas that lack common support can lead to biased inference due to imbalance or inappropriate model extrapolation. Bayesian Additive Regression Tree (BART) model with flexible functional form has many advantages over other causal inference strategies. Next section, you will learn how BART method deal with the problem of lack of overlap.

