---
title: "Balance"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(ggpubr)
library(tidyverse)
library(shiny)
library(bartCause)
library(ggforce)
library(gridExtra)
library(cowplot)
library(ggthemes) 
library(plotly)
```

```{r, echo=FALSE}
plot_scatterplot <- function(df){
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  linetypes <- c('Mean Difference' = 'solid', 'True SATE' = 'dashed')
  ggplot(df) + 
    geom_point( aes(x = X, y = Y, color = Z ))+ 
    xlab("X") + ylab("Observed Outcome") + 
    labs(title = "Distribution of Observed Outcomes") + 
    geom_hline(aes(yintercept = mean(Y[Z == 1]), color = '1', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y[Z == 0]), color = '0', linetype = 'Mean Difference')) +
    geom_hline(aes(yintercept = mean(Y1), color = '1', linetype = 'True SATE')) +
    geom_hline(aes(yintercept = mean(Y0), color = '0', linetype = 'True SATE')) +
    scale_color_manual("Treated?", values = colors) + 
    scale_linetype_manual("Linetype", values = linetypes) +
    theme_minimal(base_line_size = 0.4) + 
    theme(plot.title = element_text(size = 8), 
          legend.position="bottom",
          text=element_text(family="serif", size=8)) + 
    guides(color = guide_legend(order = 1), fill = guide_legend(order = 2))
    
}

plot_marginal_distribution <- function(df) {
  df$Z <- as.factor(df$Z)
  colors <- c('0' = 'blue', '1' = 'red')
  ggplot(df, aes(x = X, fill = Z)) + 
    theme_minimal(base_line_size = 0.4) +
    geom_histogram(bins = 60, alpha = 0.4, position = "identity") +
    geom_vline(data = df[which(df$Z == 0),], aes(xintercept= mean(X), color="0"), size=1) + 
    geom_vline(data = df[which(df$Z == 1),], aes(xintercept= mean(X), color="1"), size=1) +
    scale_color_manual(name = "Treated?", values = colors) + 
    scale_fill_manual(name = "Treated?", values = colors) +
    labs(title = "Distribution of Covariate X", y = "Frequency") +
    theme(legend.position="bottom",
          plot.title = element_text(size = 8), 
          text=element_text(family="serif", size=8)) 
    
}
```

```{r, echo=FALSE}
animation <- function(df_balanced, xaxis, yaxis){
  p <- ggplot(data = df_balanced) + geom_histogram(bins = 60, aes(x = X, color = Z, fill = Z), alpha = 0.5)
  get_hist <- function(p) {
    d <- ggplot_build(p)$data[[1]]
    data.frame(x = d$x, xmin = d$xmin, xmax = d$xmax, y = d$y, mids = (d$xmin + d$xmax)/2)
  }
  hist = get_hist(p)
  hist_group0 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  hist_group1 <- data.frame(lower = hist$xmin,
                            upper = hist$xmax,
                            count =0,
                            id = 1:length(hist),
                            x = hist$mids)
  
  idx_group0 <- which(df_balanced$Z == 0)
  idx_group1 <- which(df_balanced$Z == 1)
  
  ## group 0
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group0_first <- sample(idx_group0, 20)
  idx_group0 <- idx_group0[!idx_group0 %in% idx_group0_first]
  # the second batch to move
  idx_group0_second <- sample(idx_group0, floor(length(idx_group0)/2))
  # the third batch to move
  idx_group0_third <- idx_group0[!idx_group0 %in% idx_group0_second]
  
  # generate the dataframe for histogram animation
  # stage 3
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_first) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage3 <- hist_group0
  hist_group0_stage3$stage <- 3
  
  # stage 4
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_second) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage4 <- hist_group0
  hist_group0_stage4$stage <- 4
  
  
  # stage 5
  for (i in 1:nrow(hist_group0)) {
    for (j in idx_group0_third) {
      if(df_balanced$X[j] > hist_group0$lower[i] & df_balanced$X[j] <= hist_group0$upper[i]){
        hist_group0$count[i] = hist_group0$count[i] + 1
        next
      }
    }
  }
  hist_group0_stage5 <- hist_group0
  hist_group0_stage5$stage <- 5
  
  hist_group0_stage6 <- hist_group0_stage5
  hist_group0_stage6$stage <- 6
  hist_group0_stage7 <- hist_group0_stage5
  hist_group0_stage7$stage <- 7
  hist_group0_stage8 <- hist_group0_stage5
  hist_group0_stage8$stage <- 8
  
  hist_group0_stage1 <- hist_group0_stage5
  hist_group0_stage1$count <- 0
  hist_group0_stage1$stage <- 1
  
  hist_group0_stage2 <- hist_group0_stage5
  hist_group0_stage2$count <- 0
  hist_group0_stage2$stage <- 2
  
  histogram_group0 <- rbind(hist_group0_stage1, hist_group0_stage2, hist_group0_stage3, hist_group0_stage4, 
                            hist_group0_stage5, hist_group0_stage6, hist_group0_stage7, hist_group0_stage8)
  
  ## group 1
  # generate the indexes for scatterplot animation
  # the first batch to move
  idx_group1_first <- sample(idx_group1, floor(length(idx_group1)/3))
  idx_group1 <- idx_group1[!idx_group1 %in% idx_group1_first]
  # the second batch to move
  idx_group1_second <- sample(idx_group1, floor(length(idx_group1)/2))
  # the third batch to move
  idx_group1_third <- idx_group1[!idx_group1 %in% idx_group1_second]
  
  # generate the dataframe for histogram animation
  # stage 6
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_first) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage6 <- hist_group1
  hist_group1_stage6$stage <- 6
  
  # stage 7
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_second) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage7 <- hist_group1
  hist_group1_stage7$stage <- 7
  
  # stage 8
  for (i in 1:nrow(hist_group1)) {
    for (j in idx_group1_third) {
      if(df_balanced$X[j] > hist_group1$lower[i] & df_balanced$X[j] <= hist_group1$upper[i]){
        hist_group1$count[i] = hist_group1$count[i] + 1
        next
      }
    }
  }
  hist_group1_stage8 <- hist_group1
  hist_group1_stage8$stage <- 8
  
  
  hist_group1_stage1 <- hist_group1_stage8
  hist_group1_stage1$count <- 0
  hist_group1_stage1$stage <- 1
  
  hist_group1_stage2 <- hist_group1_stage8
  hist_group1_stage2$count <- 0
  hist_group1_stage2$stage <- 2
  
  hist_group1_stage3 <- hist_group1_stage8
  hist_group1_stage3$count <- 0
  hist_group1_stage3$stage <- 3
  
  hist_group1_stage4 <- hist_group1_stage8
  hist_group1_stage4$count <- 0
  hist_group1_stage4$stage <- 4
  
  hist_group1_stage5 <- hist_group1_stage8
  hist_group1_stage5$count <- 0
  hist_group1_stage5$stage <- 5
  
  histogram_group1 <- rbind(hist_group1_stage1, hist_group1_stage2, hist_group1_stage3, hist_group1_stage4, 
                            hist_group1_stage5, hist_group1_stage6, hist_group1_stage7, hist_group1_stage8)
  
  
  # generate the dataframe for scatterplot animation
  df_balanced_stage1 <- df_balanced  
  df_balanced_stage1$stage <- 1
  df_balanced_stage1$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage2 <- df_balanced  
  df_balanced_stage2$Y[idx_group0_first] <- sapply(df_balanced_stage2$Y[idx_group0_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage2$stage <- 2
  df_balanced_stage2$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage3 <- df_balanced  
  df_balanced_stage3$Y[idx_group0_first] <- 0
  df_balanced_stage3$Y[idx_group0_second] <- sapply(df_balanced_stage2$Y[idx_group0_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage3$stage <- 3
  df_balanced_stage3$id <- 1:nrow(df_balanced)
  
  df_balanced_stage4 <- df_balanced  
  df_balanced_stage4$Y[idx_group0_first] <- 0
  df_balanced_stage4$Y[idx_group0_second] <- 0
  df_balanced_stage4$Y[idx_group0_third] <- sapply(df_balanced_stage2$Y[idx_group0_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage4$stage <- 4
  df_balanced_stage4$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage5 <- df_balanced  
  df_balanced_stage5$Y[idx_group0_first] <- 0
  df_balanced_stage5$Y[idx_group0_second] <- 0
  df_balanced_stage5$Y[idx_group0_third] <- 0
  df_balanced_stage5$Y[idx_group1_first] <- sapply(df_balanced_stage2$Y[idx_group1_first], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage5$stage <- 5
  df_balanced_stage5$id <- 1:nrow(df_balanced)
  
  df_balanced_stage6 <- df_balanced  
  df_balanced_stage6$Y[idx_group0_first] <- 0
  df_balanced_stage6$Y[idx_group0_second] <- 0
  df_balanced_stage6$Y[idx_group0_third] <- 0
  df_balanced_stage6$Y[idx_group1_first] <- 0
  df_balanced_stage6$Y[idx_group1_second] <- sapply(df_balanced_stage2$Y[idx_group1_second], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage6$stage <- 6
  df_balanced_stage6$id <- 1:nrow(df_balanced)
  
  df_balanced_stage7 <- df_balanced  
  df_balanced_stage7$Y[idx_group0_first] <- 0
  df_balanced_stage7$Y[idx_group0_second] <- 0
  df_balanced_stage7$Y[idx_group0_third] <- 0
  df_balanced_stage7$Y[idx_group1_first] <- 0
  df_balanced_stage7$Y[idx_group1_second] <- 0
  df_balanced_stage7$Y[idx_group1_third] <- sapply(df_balanced_stage2$Y[idx_group1_third], function(x) sample(seq(0, x, by = 0.5), 1))
  df_balanced_stage7$stage <- 7
  df_balanced_stage7$id <- 1:nrow(df_balanced)
  
  
  df_balanced_stage8 <- df_balanced  
  df_balanced_stage8$Y[idx_group0_first] <- 0
  df_balanced_stage8$Y[idx_group0_second] <- 0
  df_balanced_stage8$Y[idx_group0_third] <- 0
  df_balanced_stage8$Y[idx_group1_first] <- 0
  df_balanced_stage8$Y[idx_group1_second] <- 0
  df_balanced_stage8$Y[idx_group1_third] <- 0
  df_balanced_stage8$stage <- 8
  df_balanced_stage8$id <- 1:nrow(df_balanced)
  
  
  df <- rbind(df_balanced_stage1, df_balanced_stage2, df_balanced_stage3, df_balanced_stage4, 
              df_balanced_stage5, df_balanced_stage6, df_balanced_stage7, df_balanced_stage8)
  df$Z <- as.factor(df$Z)
  
  plot_ly() %>% 
    add_markers(data=df, x = ~X, y = ~Y, color = ~Z,colors = c("blue", "red"),  frame = ~stage, ids = ~id) %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group0, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'blue'
              ),  opacity = 0.4,name = '0') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_trace(data = histogram_group1, x = ~x,
              y = ~count, base = 0,
              type = 'bar', frame = ~stage, 
              marker = list(
                color = 'red'
              ),opacity = 0.4,name = '1') %>%
    animation_opts(frame = 2000, transition = 1999, redraw = FALSE) %>% 
    add_segments(x = min(df$X) - xaxis, xend = min(df$X) - xaxis, y = 0, yend = yaxis, 
                 size = I(1), line=list(color='black'), showlegend = F) %>% 
    layout(
      # title = 'Distribution of Observed Outcomes',
      font=list(
        family='Times New Roman',
        size=12),
      barmode = 'overlay', bargap = 0,
      xaxis = list(
        title = "",
        tick = list(
          size = 14,
          color = 'rgb(107, 107, 107)')))
}
```

In a **completely randomized design**, the treatment assignment is a random variable that is independent of all variables x that occur before treatment assignment. Under repeated randomizations, there will be no differences, on average, in the pre-treatment covariates, comparing treatment and control groups. The plot shows balance in the covariate $X$ between treatment and control groups. The difference in means is an unbiased estimate of the treatment effect under a completely randomized design, and for this particular realization the estimate, 86.59 - 81.53 =  5.05, is close to the true sample average treatment effect 5.00. 

```{r, echo=FALSE}
dgp_radomized_balanced <- function (n, beta_1, seed) {
  set.seed(seed) 
  pre_quiz <- rnorm(n = n, mean = 65, sd = 3)
  beta_0 <- 10
  tau <- 5
  Z <- sample(x = c(0,1), size = n, replace = TRUE)
  Y_0 <- beta_0 + beta_1 * pre_quiz + 0 + rnorm(n) 
  Y_1 <- beta_0 + beta_1 * pre_quiz + tau + rnorm(n) 
  Y <- ifelse(Z == 1, Y_1, Y_0)
  output <- data.frame(Z, pre_quiz, Y_0, Y_1, Y)
  colnames(output) <- c( "Z", "X", "Y0", "Y1", "Y")
  return (output) 
}
df_balanced <- dgp_radomized_balanced(n = 1000, beta_1 = 1.1, seed = 123)
```

```{r, warning=FALSE, echo=FALSE}
animation(df_balanced, xaxis = 1.5, yaxis = 100)
```

```{r, echo=FALSE}
SATE_meandiff <- paste('Mean Difference:', round(mean(df_balanced[df_balanced$Z == 1, ]$Y) - mean(df_balanced[df_balanced$Z == 0, ]$Y), 2))
SATE <- paste('True SATE:', round(mean(df_balanced$Y1 - df_balanced$Y0), 2))
```

```{r, echo=FALSE}
p1 <- plot_scatterplot(df_balanced) +
  annotate(
    "text", label = SATE_meandiff,
    x = 60, y = 95, size = 3, colour = "red"
  ) + 
  annotate(
    "text", label = SATE,
    x = 60, y = 94, size = 3, colour = "red"
  )
p1_x_hist <- plot_marginal_distribution(df_balanced)
plot_grid(p1,  p1_x_hist, ncol = 2, nrow = 1)
```


In an **observational study**, there can be systematic differences between groups of units that receive different treatments with respect to key covariates, x, that can affect the outcome, y. Such covariates that are associated with the treatment and the potential outcomes are typically called confounders or confounding covariates because if we observe differences in average outcomes across these groups, we can’t separately attribute these differences to the treatment or the confounders—the effect of the treatment is thus “confounded” by these variables.

Imbalance with measured confounders occurs when the distributions of confounders differ for the treatment and control groups. This could manifest, for instance, as differences in means or standard deviations of a covariate between treatment and control groups. More generally, any differences in covariate distributions across groups can be referred to as lack of balance across groups. When treatment and control groups suffer from imbalance, the simple comparison of group averages, $\bar{y}_1− \bar{y}_0$, is not, in general, a good estimate of the average treatment effect. In the example below, the mean difference, 0.14, is far from the underlying and unobservable sample average treatment effect (SATE) of -0.5.


```{r, echo=FALSE}
dgp_overlap_unbalance <- function(n, seed){
  set.seed(seed)
  x <- runif(n,min = 10, max = 11)
  p_z <- pmin(pmax(x-10, 0.1), 0.8) # z is dependent on x
  z <- rbinom(n = n, size = 1, prob = p_z)
  y0 <- 2 * x
  y1 <- y0 - 0.5
  y <- ifelse(z==0, y0, y1) + 0.3*rnorm(n)
  return(data.frame(X=x,Z=z,Y0 = y0, Y1 = y1, Y=y))
}
df_overlap_unbalance <- dgp_overlap_unbalance(1000, seed = 1234)
```

```{r, warning=FALSE, echo=FALSE}
animation(df_overlap_unbalance, xaxis = 0.05, yaxis = 25)
```

```{r, echo=FALSE}
SATE_meandiff <- paste('Mean Difference:', round(mean(df_overlap_unbalance$Y[df_overlap_unbalance$Z == 1]) -
  mean(df_overlap_unbalance$Y[df_overlap_unbalance$Z == 0]), 2))
SATE <- paste('True SATE:', round(mean(df_overlap_unbalance$Y1 - df_overlap_unbalance$Y0), 2))
```

```{r, echo=FALSE}
p2 <- plot_scatterplot(df_overlap_unbalance) +
  annotate(
    "text", label = SATE_meandiff,
    x = 10.2, y = 22, size = 3, colour = "red"
  ) + 
  annotate(
    "text", label = SATE,
    x = 10.2, y = 21.9, size = 3, colour = "red"
  )
p2_x_hist <- plot_marginal_distribution(df_overlap_unbalance)
plot_grid(p2,  p2_x_hist, ncol = 2, nrow = 1)
```

As the pre-treatment variable is imbalanced between the two groups, the mean difference is biased. Another approach to estimate the average treatment effect is to fit a linear regression of the outcome $Y$ on the treatment variable $Z$ and the confounder $X$ which can help us adjust for systematically unbalanced characteristics across groups. 

```{r, echo=FALSE}
inputPanel(selectInput("select_lm_result", h4("Select the result from the regression to show: "), 
                          choices = c("Only show the coefficient estimated for the treatment effect" = "coefficient", 
                                      "Show the full regression output" = "full_output"), 
                          selected = "coefficient"),
           conditionalPanel("input.select_lm_result === 'full_output'",
                            verbatimTextOutput('lm_result')),
           conditionalPanel("input.select_lm_result === 'coefficient'",
                            verbatimTextOutput('coef')))

df_overlap_unbalance$Z <- as.factor(df_overlap_unbalance$Z)           
model <- lm(Y~Z+X, data = df_overlap_unbalance)

output$lm_result <- renderPrint({summary(model)})
output$coef <- renderPrint({summary(model)$coefficients[,1]["Z1"]})
```

In this case linear regression has done well in identifying the correct ATE - which is good, but the data generating process was specifically designed to meet the assumptions. Let's look at a case where it might fail.

```{r, echo=FALSE}
dgp_nonoverlap_balance <- function(n, seed){
  set.seed(seed)
  num_1 <- floor(n/2)
  num_0 <- n - num_1
  half <- floor(num_1/2)
  x1 <- c(rnorm(half, mean = -2, sd = 1), rnorm(num_1-half, mean = 2, sd = 1))
  x0 <- rnorm(num_0, mean = 0, sd = 1)
  x <- c(x1, x0)
  # use the boundaries of bars in the histogram to ensure no overlap between the two groups in the histogram
  p_z <- ifelse(x<1.14896350&x>-1.14896350, 1, 0) 
  z <- rbinom(n = n, size = 1, prob = p_z)
  y0 =  ifelse(x<1&x>-1, 30 + x - 0.2*x^2 - 10 + rnorm(n), 30 + x - 0.2*x^2 + rnorm(n))
  y1 =  30 + x + 0.4*x^2 + 5 + rnorm(n)
  y = ifelse(z == 0, y0, y1)
  return(data.frame(X=x,Z=z,Y0 = y0, Y1 = y1, Y=y))
}
df_nonoverlap_balance <- dgp_nonoverlap_balance(1000, 1234)
```

```{r, warning=FALSE, echo=FALSE}
animation(df_nonoverlap_balance, xaxis = 0.5, yaxis = 50)
```

```{r, echo=FALSE}
SATE_meandiff <- paste('Mean Difference:', round(mean(df_nonoverlap_balance$Y[df_nonoverlap_balance$Z == 1]) -
  mean(df_nonoverlap_balance$Y[df_nonoverlap_balance$Z == 0]), 2))
SATE <- paste('True SATE:', round(mean(df_nonoverlap_balance$Y1 - df_nonoverlap_balance$Y0), 2))
```

```{r, echo=FALSE}
p3 <- plot_scatterplot(df_nonoverlap_balance) +
  annotate(
    "text", label = SATE_meandiff,
    x = -3, y = 40, size = 3, colour = "red"
  ) + 
  annotate(
    "text", label = SATE,
    x = -3, y = 39, size = 3, colour = "red"
  )
p3_x_hist <- plot_marginal_distribution(df_nonoverlap_balance)
plot_grid(p3,  p3_x_hist, ncol = 2, nrow = 1)
```

The pre-treatment variable $X$ is balanced in terms of group mean between the two groups, but the mean difference and linear regression both fail to identify the correct ATE. This is due to the problem of lack of overlap, which is introduced in the next section.

```{r, echo=FALSE}
df_nonoverlap_balance$Z <- as.factor(df_nonoverlap_balance$Z)
model_fail <- lm(Y~X+Z, data = df_nonoverlap_balance)

inputPanel(selectInput("select_lm_fail_result", h4("Select the result from the regression to show: "), 
                       choices = c("Only show the coefficient estimated for the treatment effect" = "lm_fail_coefficient", 
                                   "Show the full regression output" = "lm_fail_full_output"), 
                       selected = "lm_fail_coefficient"),
           conditionalPanel("input.select_lm_fail_result === 'lm_fail_full_output'",
                            verbatimTextOutput('balance_lm_fail_result')),
           conditionalPanel("input.select_lm_fail_result === 'lm_fail_coefficient'",
                            verbatimTextOutput('balance_lm_fail_coef')))

output$balance_lm_fail_result <- renderPrint({summary(model_fail)})
output$balance_lm_fail_coef <- renderPrint({round(summary(model_fail)$coefficients[,1]["Z1"],2)})


```


