---
title: "Untitled"
output: html_document
runtime: shiny
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
library(bartCause)
library(aciccomp2016)
library(tidyverse)
```

The recommended approach to variable selection is to include any pre-treatment covariate that could be predictive of the outcome variable. Although there are caveats to this advice they are special cases and will be discussed in a subsequent module. This recommendation means that even if there are many pre-treatment variables, they should all be included in the analysis! **When using thinkCausal, concern about including too many pre-treatment predictors is not a valid reason to exclude variables from a causal analysis.**


Introductory statistics courses often recomend there should be 10 observations for each predictor variable. Under this approach a statistical model fit to a dataset of 100 observations should have no more than 10 predictor variables. 


```{r}
# get dat
# data("input_2016")
# dat <- input_2016
# n <- 100
# dat <- fastDummies::dummy_cols(dat, select_columns = c('x_2', 'x_21', 'x_24'), remove_selected_columns = TRUE)
# names(dat) <- paste0('x_', 1:length(dat))

```

```{r}
# counter <- reactiveVal(value = 0)
# 
# draw_sample1 <- reactive({
#   
#   index <- sample(1:nrow(dat), n)
#   X <- dat[index, ]
#   oracle <- dgp_2016(input_2016, 4, 1)
#   y <- oracle$y[index]
#   z <- oracle$z[index]
#   ice <- oracle$y.1[index] - oracle$y.0[index]
#   att <- mean(ice[z == 1])
# 
#   predictors <- n/10 
#   adjust_prediction_y <- names(tail(lm(y ~ 0 + ., X)$coeff, predictors))
#   adjust_prediction_z <- names(tail(glm(z ~ 0 + .,X, family = binomial)$coeff, predictors))
# 
#   p.z <- vector()
#   for (i in 1:length(dat)) {
#     coef <- glm(z ~ 0 + X[,i],X, family = binomial)$coef
#     se <- sqrt(diag(vcov(glm(z ~ 0 + X[,i], X,family = binomial))))
#     p.z[i] <- 2*(1-pt(abs(coef/se),99))
#   
#   }
# 
#   p.z[is.na(p.z)] <- F
#   coefs <- lm(y ~ 0 + ., X)$coeff
#   coefs <- coefs[p.z <.05][order(abs(coefs[p.z <.05]))] %>% na.omit()
#   mix <- names(tail(coefs, predictors))
#   
#   adjust_prediction_y <- names(coefs[1:predictors])
#   allX <- summary(bartc(y, z, X, estimand = 'att'), target = 'sate')
#   top10z <- summary(bartc(y, z, X[,adjust_prediction_z], estimand = 'att'), target = 'sate')
#   top10y <- summary(bartc(y, z, X[,adjust_prediction_y], estimand = 'att'), target = 'sate')
#   mix_model <- summary(bartc(y, z, X[,mix], estimand = 'att'), target = 'sate')
#   random <- summary(bartc(y, z, X[,sample(1:length(dat), predictors)], estimand = 'att'), target = 'sate')
#   
#   
#   results <- rbind(allX$estimates, top10y$estimates, top10z$estimates, mix_model$estimates, random$estimates)
#   results$model <-
#     c(
#       'all covariates',
#       'top 10 predictors of y',
#       'top 10 predictors of z',
#       'top 10 predictors of y that are significantly associated with z', 
#       '10 randomly selected covaraites'
#     )
#   
#   out <- list(results, att)
#   
#   return(out)
# 
# })


```



```{r}
# set.seed(21)
# x <- rnorm(10)
# 
# y <- 20 + 5*x + rnorm(10, 0, 3)
# a1 <- rbinom(10, 1, .5)
# a2 <- rbinom(10, 1, .5)
# a3 <- rbinom(10, 1, .5)
# a4 <- rbinom(10, 1, .5)
# a5 <- rbinom(10, 1, .5)
# a6 <- rbinom(10, 1, .5)
# a7 <- rbinom(10, 1, .5)
# a8 <- rbinom(10, 1, .5)
# a9 <- rbinom(10, 1, .5)
# 
# overfit <- data.frame(a1, a2, a3, a4, a5, a6, a7, x, y)
# 
# check <- lm(y ~ ., overfit)
# tibble(overfit, prediction = check$fitted.values) %>% 
# ggplot(aes(x, y)) + 
#   geom_point() + 
#   geom_line(aes(x, prediction))
# 
# x <- rnorm(10)
# 
# new_y <- 20 + 5*x + rnorm(10, 0, 3)
# a1 <- rbinom(10, 1, .5)
# a2 <- rbinom(10, 1, .5)
# a3 <- rbinom(10, 1, .5)
# a4 <- rbinom(10, 1, .5)
# a5 <- rbinom(10, 1, .5)
# a6 <- rbinom(10, 1, .5)
# a7 <- rbinom(10, 1, .5)
# a8 <- rbinom(10, 1, .5)
# a9 <- rbinom(10, 1, .5)
# new_dat <- data.frame(a1, a2, a3, a4, a5, a6, a7, x)
# new_prediction <- predict(check, newdata = new_dat)
# summary(check)
# tibble(prediction = new_prediction, new_y, new_x = x) %>% 
# ggplot(aes(new_x, new_y)) + 
#   geom_point() + 
#   geom_line(aes(new_x, prediction))
# new_preds <- predict(check, new_dat)
```


```{r}
set.seed(21)
x <- rnorm(8)
a <- sample(c('black', 'red', 'green', 'blue', 'teel','purple', 'yellow', 'grey'), 8)
a <- factor(a, levels = c('black', 'red', 'green', 'blue', 'teel','purple', 'yellow', 'grey'))
overfit_dat <- tibble(x,a)
y <- 20 + x*5 + rnorm(8, 0, 1)

overfit_dat <- data.frame(y, overfit_dat)

check <- lm(y ~ 0 + ., overfit_dat)

check
ggplot(overfit_dat, aes(x, y, col = a)) + 
  geom_point(size = 3) + 
  geom_abline(intercept = 0, slope = check$coefficients[1]) + 
  geom_abline(intercept = check$coefficients[2], slope = check$coefficients[1]) + 
    geom_abline(intercept = check$coefficients[3], slope = check$coefficients[1]) + 
      geom_abline(intercept = check$coefficients[4], slope = check$coefficients[1]) +
        geom_abline(intercept = check$coefficients[5], slope = check$coefficients[1]) + 
        geom_abline(intercept = check$coefficients[6], slope = check$coefficients[1]) + 
        geom_abline(intercept = check$coefficients[7], slope = check$coefficients[1])  + 
          geom_abline(intercept = check$coefficients[8], slope = check$coefficients[1]) + 
          scale_color_manual(values = 1:8) + 
  coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(-10, 90))


x <- rnorm(8)
a <- sample(c('black', 'red', 'green', 'blue', 'teel','purple', 'yellow', 'grey'), 8)
a <- factor(a, levels = c('black', 'red', 'green', 'blue', 'teel','purple', 'yellow', 'grey'))
new_y <- 20 + x*5 + rnorm(8, 0, 1)

predicted_y <- predict(check, new_dat)
new_dat <- tibble(x,a, new_y, predicted_y)


ggplot(new_dat, aes(x, new_y, col = a)) + 
  geom_point(size = 3) + 
  geom_point(size = 3, shape = 1, aes(x, predicted_y)) + 
  geom_abline(intercept = 0, slope = check$coefficients[1]) + 
  geom_abline(intercept = check$coefficients[2], slope = check$coefficients[1]) + 
    geom_abline(intercept = check$coefficients[3], slope = check$coefficients[1]) + 
      geom_abline(intercept = check$coefficients[4], slope = check$coefficients[1]) +
        geom_abline(intercept = check$coefficients[5], slope = check$coefficients[1]) + 
        geom_abline(intercept = check$coefficients[6], slope = check$coefficients[1]) + 
        geom_abline(intercept = check$coefficients[7], slope = check$coefficients[1])  + 
          geom_abline(intercept = check$coefficients[8], slope = check$coefficients[1]) + 
          scale_color_manual(values = 1:8) + 
    coord_cartesian(xlim = c(-2.5, 2.5), ylim = c(-10, 90))



predicted_y;new_y
```

